---
title: "Bayesian Analysis of Food Sharing Behavior in Rats"
subtitle: "A Replication of Wan et al. (2021), Frontiers in Psychology"
author: "Haoran (Matt) Wan"
date: "today"
format: 
  html:
    toc: true
    code-fold: false
    self-contained: true
    theme: cosmo
    mainfont: "Garamond"
execute:
  warning: false
  message: false
engine: knitr
---

## Project Objective

This document provides the complete R code to replicate the Bayesian statistical analyses from the publication:

> Wan, H., Kirkman, C. F., Jensen, G., & Hackenberg, T. D. (2021). Failure to find altruistic food sharing in rats. *Frontiers in Psychology*, *12*, Article 696025. https://doi.org/10.3389/fpsyg.2021.696025

The study's primary goal was to conduct a systematic replication to evaluate prior claims of altruistic food sharing in rats. The experiment involved rats choosing between food rewards and social interaction under varying conditions of food availability and motivation.

The data and original analysis scripts are available in the Supplementary Material of the original publication at the **[publisher's website](https://www.frontiersin.org/articles/10.3389/fpsyg.2021.696025/full#supplementary-material)**.

## Analytical Approach

The analysis relies on **Bayesian hierarchical generalized linear models** implemented in Stan via the `cmdstanr` package. This approach is well-suited for modeling behavioral data with repeated measures from a small number of subjects, providing robust estimates and full posterior inference. Specifically, we fit:

1.  A **hierarchical binomial logistic regression** to model the choice proportion between food and social interaction.
2.  A **hierarchical negative binomial regression** to model the counts of food and social responses, accounting for potential overdispersion in the count data.
3.  A **hierarchical negative binomial regression** to model the counts of food pellets consumed versus shared.

## Analysis Workflow

1.  **Setup**: Load necessary R packages and the raw experimental data. Prepare data for Stan modeling.
2.  **Model 1 (Choice Proportion)**: Fit and summarize the logistic regression model.
3.  **Model 2 (Response Counts)**: Fit and summarize the negative binomial model for total food/social responses.
4.  **Model 3 (Pellet Allocation)**: Fit and summarize the negative binomial model for consumed/shared pellets.

---

## 1. Setup: Load Packages and Prepare Data

This section loads the required R packages and prepares the raw data file for analysis with the Stan models.

```{r setup}
# --- 1.1 Load Required Packages ---
library(cmdstanr) # Interface for running Stan models
library(readr)    # For reading CSV files efficiently

# --- 1.2 Load and Prepare Data ---
# Load the raw dataset from the CSV file provided in the supplementary materials.
raw_data <- read_csv("Raw_data.csv")

# Prepare the data for Stan modeling:
# Convert the 'Rat' identifier into a sequential integer factor ('Subject').
# Stan models typically require integer indices for group-level parameters.
analysis_data <- raw_data
analysis_data$Subject <- as.integer(as.factor(analysis_data$Rat))
```

## 2. Model 1: Choice Proportion (Food vs. Social)

The first analysis models the rats' **preference** between the food and social options. We use a **Bayesian hierarchical binomial logistic regression** to estimate the probability of choosing the food lever over the social lever in each experimental condition.

**Model Structure:**

* **Outcome:** The number of `Food` choices out of the total `Social + Food` choices in each session.
* **Likelihood:** Binomial distribution with a logit link function.
* **Predictors:** Experimental condition (as a categorical variable).
* **Hierarchical Component:** The model includes varying intercepts for each `Subject` within each `Condition`. This accounts for individual differences in preference and how those differences might change across conditions. The `gamma` parameters represent the population-level average log-odds of choosing food in each condition.

**Data Subsetting:** As noted in the publication, Condition 5 is excluded from this analysis because only the food lever was active (no social choice was available), making preference undefined for that condition.

### Stan Model Code and Execution

The R chunk below performs the following steps:

1.  Defines the Stan model code for the hierarchical logistic regression.
2.  Writes the code to a temporary `.stan` file.
3.  Compiles the Stan model using `cmdstanr`.
4.  Prepares the data list required by the Stan model, including filtering out Condition 5 and re-indexing the conditions.
5.  Runs the MCMC sampler to draw samples from the posterior distribution.
6.  Prints a summary table focused on the population-level coefficients (`gamma`), which represent the average preference in each condition on the log-odds scale.

```{r run_pair_model}
# --- 2.1 Stan Model Definition (Binomial Logistic Regression) ---

# Define the Stan code for the multilevel binomial logistic regression model.
# This model estimates the probability of choosing food, allowing the intercept (preference)
# to vary by subject and condition.
choice_model_code <- "
data {
  int<lower=1> N;                // Number of observations (sessions)
  int<lower=1> S;                // Number of subjects (rats)
  array[N] int<lower=1> subID;   // Subject ID for each observation
  int<lower=1> C;                // Number of conditions
  array[N] int<lower=1> condID;  // Condition ID for each observation
  array[N] int<lower=0> choice;  // Total choices (Food + Social) in session
  array[N] int<lower=0> food;    // Number of food choices in session
}

transformed data {
  // Vector of ones for simpler matrix math in beta calculation
  vector[S] u = rep_vector(1.0, S); 
}

parameters {
  // Parameters for the hierarchical structure (non-centered parameterization)
  matrix[C, S] z;                   // Standardized subject deviations
  cholesky_factor_corr[C] L_Omega;  // Cholesky factor of the correlation matrix for varying effects across conditions
  vector<lower=0>[C] tau;           // Scale parameters for varying effects (subject SDs per condition)
  row_vector[C] gamma;              // Population-level intercepts (mean log-odds per condition)
}

transformed parameters {
  // Calculate subject-specific intercepts (beta) for each condition
  // beta = population_mean + subject_deviation
  // The complex term calculates correlated subject deviations efficiently
  matrix[S, C] beta = u * gamma + (diag_pre_multiply(tau, L_Omega) * z)';
}

model {
  // --- Priors ---
  // Weakly informative priors
  to_vector(z) ~ std_normal();         // Standard normal for standardized deviations
  L_Omega ~ lkj_corr_cholesky(2);     // LKJ prior for correlation matrix (favors weak correlations)
  tau ~ exponential(1.5);             // Exponential prior for scale parameters (SDs)
  to_vector(gamma) ~ normal(0, 1.5);  // Normal prior for population means

  // --- Likelihood ---
  // The number of food choices follows a binomial distribution,
  // where the probability is determined by the logit-transformed subject/condition intercept (mu).
  array[N] real mu;
  for (n in 1:N) {
    mu[n] = beta[subID[n], condID[n]]; // Get the appropriate intercept for this subject/condition
  }
  food ~ binomial_logit(choice, mu); 
}
"

# --- 2.2 Compile Stan Model ---
# Write the Stan code to a temporary file
stan_file_choice <- write_stan_file(choice_model_code)
# Compile the model using cmdstanr (may take a moment on first run)
choice_model_compiled <- cmdstan_model(stan_file_choice)

# --- 2.3 Prepare Data for Stan ---
# Filter out Condition 5, as it had no social choice option
analysis_rows_choice <- (analysis_data$Condition != 5)
data_subset_choice <- analysis_data[analysis_rows_choice, ]

# Re-index Condition IDs sequentially from 1 to C (required by Stan)
# Conditions 6 and 7 become 5 and 6 after removing original Condition 5.
data_subset_choice$Condition_Stan <- data_subset_choice$Condition
data_subset_choice$Condition_Stan[data_subset_choice$Condition > 5] <- 
    data_subset_choice$Condition[data_subset_choice$Condition > 5] - 1

# Create the list object containing data in the format Stan expects
stan_data_choice <- list(
  N = nrow(data_subset_choice),
  S = length(unique(data_subset_choice$Subject)),
  subID = data_subset_choice$Subject,
  C = length(unique(data_subset_choice$Condition_Stan)),
  condID = data_subset_choice$Condition_Stan,
  choice = data_subset_choice$Social + data_subset_choice$Food, # Total choices
  food = data_subset_choice$Food                                # Number of successes (food choices)
)

# --- 2.4 Run MCMC Sampler ---
# Sample from the posterior distribution using the No-U-Turn Sampler (NUTS)
choice_model_fit <- choice_model_compiled$sample(
  data = stan_data_choice,
  chains = 4,             # Number of independent Markov chains
  parallel_chains = 4,    # Run chains in parallel using available cores
  iter_warmup = 1500,     # Number of warmup (burn-in) iterations per chain
  iter_sampling = 1500,   # Number of sampling iterations per chain
  adapt_delta = 0.95,     # Target acceptance probability (higher value reduces divergent transitions)
  max_treedepth = 15,     # Maximum tree depth (limits complexity of NUTS steps)
  refresh = 0,            # Suppress iteration progress output
  show_messages = FALSE,  # Suppress Stan informational messages
  show_exceptions = FALSE # Suppress Stan exceptions (if any)
)

# --- 2.5 Summarize Results ---
# Display a summary table for the population-level parameters (gamma).
# These represent the average log-odds of choosing food in each condition (1-4, 6, 7).
cat("--- Summary of Population-Level Parameters (gamma: Log-Odds of Food Choice) ---\n")
choice_model_fit$summary(variables = "gamma")
```

## 3. Model 2: Response Counts (Food and Social)

While the first model examined the rats' **preference** (proportion of choices), this second analysis models the **absolute rate** of responding for food and social interaction in each session. This allows us to see if the experimental conditions affected the overall number of choices made, not just the distribution between the two options.

**Model Structure:**

* **Outcome:** The number of `Food` choices and `Social` choices per session (modeled simultaneously).
* **Likelihood:** **Negative Binomial distribution**. This is a generalization of the Poisson distribution that is appropriate for count data. It includes an extra parameter to account for **overdispersion**—the tendency for the variance in behavioral count data to be greater than the mean—which is common in this type of experiment.
* **Predictors:** Experimental condition and response type (Food vs. Social).
* **Hierarchical Component:** The model estimates a unique population-level log-rate (`gamma`) for each response type in each condition. It also includes varying intercepts for each `Subject`, allowing each rat's baseline response rate to differ from the population average.

### Stan Model Code and Execution

The R chunk below follows the same process as before: defining, compiling, preparing data for, and running a custom Stan model. The key difference is the use of a negative binomial likelihood to model the response counts. The summary table shows the `gamma` parameters, which represent the population-level log-rates for food choices (`gamma[1]` to `gamma[7]`) and social choices (`gamma[8]` to `gamma[14]`) in each of the seven conditions.

```{r run_resp_rate_model}
# --- 3.1 Stan Model Definition (Negative Binomial Regression) ---

# Define the Stan code for the multilevel negative binomial regression model.
# This model estimates the rate of Food and Social choices, accounting for
# overdispersion in the count data.
rate_model_code <- "
data {
  int<lower=1> N;                // Number of observations
  int<lower=1> S;                // Number of subjects
  array[N] int<lower=1> subID;   // Subject ID for each observation
  int<lower=1> C;                // Number of conditions
  array[N] int<lower=1> condID;  // Condition ID for each observation
  array[N] int<lower=0> FC;      // Food choices (counts)
  array[N] int SC;               // Social choices (counts), allows non-positive for NA handling
}

parameters {
  // Parameters for the hierarchical structure (2*C for Food and Social responses)
  matrix[2*C, S] z;                 // Standardized subject deviations
  cholesky_factor_corr[2*C] L_Omega;// Cholesky factor of the correlation matrix
  vector<lower=0>[2*C] tau;         // Scale parameters for varying effects
  row_vector[2*C] gamma;            // Population-level log-rates
  real<lower=0> overdisp;           // Overdispersion parameter for the negative binomial
}

transformed parameters {
  // Calculate subject-specific log-rates (beta) for each response type and condition
  matrix[S, 2*C] beta = rep_matrix(gamma, S) + (diag_pre_multiply(tau, L_Omega) * z)';
}

model {
  // --- Priors ---
  to_vector(z) ~ std_normal();
  L_Omega ~ lkj_corr_cholesky(2);
  tau ~ exponential(1.5);
  to_vector(gamma) ~ normal(1, 1.5); // Prior centered at 1 (exp(1) ~= 2.7 choices)
  overdisp ~ exponential(1);

  // --- Likelihood ---
  // Model food and social choices as separate negative binomial processes
  for (n in 1:N) {
    // Food choices are modeled for all conditions
    FC[n] ~ neg_binomial_2_log(beta[subID[n], condID[n]], overdisp);
    
    // Social choices are only modeled if they are not NA (passed as -1 or another negative number)
    // This correctly skips the likelihood calculation for Condition 5.
    if (SC[n] >= 0) {
      SC[n] ~ neg_binomial_2_log(beta[subID[n], condID[n] + C], overdisp);
    }
  }
}
"
# --- 3.2 Compile Stan Model ---
stan_file_rate <- write_stan_file(rate_model_code)
rate_model_compiled <- cmdstan_model(stan_file_rate)

# --- 3.3 Prepare Data for Stan ---
# The Social choice data contains NAs for Condition 5. Stan's integer type
# does not support NAs, so we replace them with -1. The Stan model code
# uses an `if` statement to skip these observations in the likelihood calculation.
stan_data_rate <- list(
  N = nrow(analysis_data),
  S = length(unique(analysis_data$Subject)),
  subID = analysis_data$Subject,
  C = length(unique(analysis_data$Condition)),
  condID = analysis_data$Condition,
  FC = analysis_data$Food,
  SC = tidyr::replace_na(analysis_data$Social, -1) # Replace NA with -1
)

# --- 3.4 Run MCMC Sampler ---
rate_model_fit <- rate_model_compiled$sample(
  data = stan_data_rate,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1500,
  iter_sampling = 1500,
  adapt_delta = 0.95,
  max_treedepth = 15,
  refresh = 0,
  show_messages = FALSE,
  show_exceptions = FALSE
)

# --- 3.5 Summarize Results ---
# Display a summary for the population-level parameters (gamma).
# gamma[1]..gamma[7] are the log-rates for Food choices in Conditions 1-7.
# gamma[8]..gamma[14] are the log-rates for Social choices in Conditions 1-7 (except 5).
cat("--- Summary of Population-Level Parameters (gamma: Log-Rates of Response) ---\n")
rate_model_fit$summary(variables = "gamma")
```

## 4. Model 3: Pellet Allocation (Consumed vs. Shared)

This final analysis addresses the central question of the study: **do rats altruistically share food?** We model the number of food pellets that were **consumed** by the focal rat versus the number that were **shared** (i.e., produced but left available for the partner rat).

This model allows us to directly quantify and test the rate of sharing behavior across the different experimental conditions, providing a direct test of the altruism hypothesis.

**Model Structure:**

* **Outcome:** The number of `Consumed_Food` pellets and `Shared_Food` pellets per session.
* **Likelihood:** **Negative Binomial distribution**, again used to correctly model overdispersed count data.
* **Predictors:** Experimental condition and pellet type (Consumed vs. Shared).
* **Hierarchical Component:** The model estimates a population-level log-rate (`gamma`) for each pellet type in each condition, with varying intercepts for each `Subject`.

### Stan Model Code and Execution

The structure of this model is nearly identical to Model 2, but the input data is now the counts of consumed and shared pellets. The summary table for `gamma` shows the population-level log-rates for consumed pellets (`gamma[1]` to `gamma[7]`) and shared pellets (`gamma[8]` to `gamma[14]`).

```{r run_intake_model}
# --- 4.1 Stan Model Definition (Negative Binomial Regression for Pellets) ---

# This Stan code is structurally identical to the previous model but is now
# intended to model the counts of Consumed vs. Shared food pellets.
pellet_model_code <- "
data {
  int<lower=1> N;
  int<lower=1> S;
  array[N] int<lower=1> subID;
  int<lower=1> C;
  array[N] int<lower=1> condID;
  array[N] int<lower=0> Consumed; // Outcome 1: Consumed pellets
  array[N] int Shared;           // Outcome 2: Shared pellets (allows -1 for NA)
}

parameters {
  matrix[2*C, S] z;
  cholesky_factor_corr[2*C] L_Omega;
  vector<lower=0>[2*C] tau;
  row_vector[2*C] gamma;
  real<lower=0> overdisp;
}

transformed parameters {
  matrix[S, 2*C] beta = rep_matrix(gamma, S) + (diag_pre_multiply(tau, L_Omega) * z)';
}

model {
  // Priors
  to_vector(z) ~ std_normal();
  L_Omega ~ lkj_corr_cholesky(2);
  tau ~ exponential(1.5);
  to_vector(gamma) ~ normal(1, 1.5);
  overdisp ~ exponential(1);

  // Likelihood
  for (n in 1:N) {
    Consumed[n] ~ neg_binomial_2_log(beta[subID[n], condID[n]], overdisp);
    // Skip likelihood for Shared pellets where it's not applicable (e.g., Condition 5)
    if (Shared[n] >= 0) {
      Shared[n] ~ neg_binomial_2_log(beta[subID[n], condID[n] + C], overdisp);
    }
  }
}
"
# --- 4.2 Compile Stan Model ---
stan_file_pellet <- write_stan_file(pellet_model_code)
pellet_model_compiled <- cmdstan_model(stan_file_pellet)

# --- 4.3 Prepare Data for Stan ---
# As before, replace NAs in the 'PelletsLeft' column with -1 for Stan compatibility.
stan_data_pellet <- list(
  N = nrow(analysis_data),
  S = length(unique(analysis_data$Subject)),
  subID = analysis_data$Subject,
  C = length(unique(analysis_data$Condition)),
  condID = analysis_data$Condition,
  Consumed = analysis_data$Food,
  Shared = tidyr::replace_na(analysis_data$PelletsLeft, -1) # Replace NA with -1
)

# --- 4.4 Run MCMC Sampler ---
pellet_model_fit <- pellet_model_compiled$sample(
  data = stan_data_pellet,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1500,
  iter_sampling = 1500,
  adapt_delta = 0.95,
  max_treedepth = 15,
  refresh = 0,
  show_messages = FALSE,
  show_exceptions = FALSE
)

# --- 4.5 Summarize Results ---
# Display a summary for the population-level parameters (gamma).
# gamma[1]..gamma[7] are the log-rates for Consumed pellets in Conditions 1-7.
# gamma[8]..gamma[14] are the log-rates for Shared pellets in Conditions 1-7.
cat("--- Summary of Population-Level Parameters (gamma: Log-Rates of Pellet Allocation) ---\n")
pellet_model_fit$summary(variables = "gamma")
```