---
title: "Bayesian Analysis of Food Sharing Behavior in Rats"
subtitle: "Reproducible R implementation of Wan et al. (2021)"
author: "Haoran Wan"
date: "2026-07-01"
format:
  html:
    toc: true
    code-fold: false
    self-contained: true
    theme: cosmo
execute:
  warning: true
  message: true
engine: knitr
---

## Scientific reference

This document is a maintained R reimplementation of the model specification reported in:

> Wan, H., Kirkman, C. F., Jensen, G., & Hackenberg, T. D. (2021). Failure to find altruistic food sharing in rats. *Frontiers in Psychology, 12*, 696025. <https://doi.org/10.3389/fpsyg.2021.696025>

The unmodified 2021 analysis files are preserved in [`archive/published-2021/`](archive/published-2021/). They are the scientific source of truth. The companion workflow that preceded the corrections in this document is preserved in [`archive/companion-pre-correction/`](archive/companion-pre-correction/) for provenance.

The published analysis defines three Bayesian multilevel generalized linear models:

1. A binomial-logistic model for food choice relative to social choice.
2. A negative-binomial model for food- and social-response counts.
3. A negative-binomial model for three pellet outcomes: the published consumed-pellet measure (`CP`), left behind (`LP`), and shared (`SP`).

The archived code labels `CP` as pellets consumed and calculates it as `Food * FoodAmmt`. This document preserves that published operationalization rather than redefining it: `CP = Food * FoodAmmt`, `LP = PelletsLeft`, and `SP = Sharing`.

## Study and inferential scope

Six female Sprague-Dawley rats were studied as three familiar pairs. The three focal rats generated the session-level observations analyzed here; their partners were available for social release and, when the procedure allowed, food sharing. Food magnitude, home-cage food access, food location, and the food-collection period varied across seven conditions.

The models estimate posterior distributions and posterior predictions. They do not implement null-hypothesis significance tests. Population-level parameters describe these three focal rats under the tested procedures; generalization beyond them should remain appropriately cautious.

| Model | Outcome | Likelihood | Population-scale interpretation |
| --- | --- | --- | --- |
| Choice | Food choices out of food + social choices | Binomial-logistic | Food-choice probability at zero subject deviation |
| Response rate | Food and social choices per session | Negative binomial | Expected choices at zero subject deviation |
| Pellet allocation | Published `CP`, `LP`, and `SP` counts | Negative binomial | Expected pellets at zero subject deviation |

## Reproducibility configuration

Run this document from the `Analysis/` directory. CmdStan is pinned in [`../config/cmdstan-version.txt`](../config/cmdstan-version.txt); execution stops if a different version is active. Four chains, 3,000 warmup iterations, 3,000 retained iterations, `adapt_delta = 0.99`, and `max_treedepth = 15` match the settings in the archived analysis. The fixed seeds support deterministic re-execution of this workflow but cannot reproduce the exact 2021 random draws across different Stan interfaces and versions.

```{r setup}
library(cmdstanr)
library(readr)

PINNED_CMDSTAN_VERSION <- trimws(readLines("../config/cmdstan-version.txt", warn = FALSE)[1])
INSTALLED_CMDSTAN_VERSION <- tryCatch(
  as.character(cmdstan_version()),
  error = function(e) NA_character_
)

if (is.na(INSTALLED_CMDSTAN_VERSION) ||
    INSTALLED_CMDSTAN_VERSION != PINNED_CMDSTAN_VERSION) {
  stop(
    sprintf(
      paste0(
        "CmdStan %s is required; active version is %s. ",
        "Install the pinned version before rendering."
      ),
      PINNED_CMDSTAN_VERSION,
      ifelse(is.na(INSTALLED_CMDSTAN_VERSION), "not available", INSTALLED_CMDSTAN_VERSION)
    )
  )
}

MODEL_SEEDS <- c(
  choice = 20210622L,
  response_rate = 20210623L,
  pellet = 20210624L
)

SAMPLING_CONFIG <- list(
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 3000,
  iter_sampling = 3000,
  adapt_delta = 0.99,
  max_treedepth = 15,
  refresh = 500
)

sample_maintained_model <- function(model, data, seed) {
  model$sample(
    data = data,
    seed = seed,
    chains = SAMPLING_CONFIG$chains,
    parallel_chains = SAMPLING_CONFIG$parallel_chains,
    iter_warmup = SAMPLING_CONFIG$iter_warmup,
    iter_sampling = SAMPLING_CONFIG$iter_sampling,
    adapt_delta = SAMPLING_CONFIG$adapt_delta,
    max_treedepth = SAMPLING_CONFIG$max_treedepth,
    refresh = SAMPLING_CONFIG$refresh
  )
}

report_diagnostics <- function(fit, model_name) {
  cat(sprintf("\n=== %s: sampler diagnostics ===\n", model_name))
  diagnostics <- fit$diagnostic_summary(quiet = FALSE)
  print(diagnostics)

  cat(sprintf("\n=== %s: CmdStan diagnose ===\n", model_name))
  fit$cmdstan_diagnose()

  summary <- fit$summary()
  cat(sprintf("\n=== %s: parameter diagnostics ===\n", model_name))
  print(
    data.frame(
      max_rhat = max(summary$rhat, na.rm = TRUE),
      min_ess_bulk = min(summary$ess_bulk, na.rm = TRUE),
      min_ess_tail = min(summary$ess_tail, na.rm = TRUE)
    ),
    row.names = FALSE
  )

  invisible(list(sampler = diagnostics, parameters = summary))
}

summarize_gamma <- function(fit, labels, inverse_link) {
  draws <- fit$draws(variables = "gamma", format = "matrix")
  gamma_columns <- grepl("^gamma\\[", colnames(draws))
  natural_draws <- inverse_link(draws[, gamma_columns, drop = FALSE])

  stopifnot(ncol(natural_draws) == length(labels))
  intervals <- t(apply(
    natural_draws,
    2,
    quantile,
    probs = c(0.025, 0.10, 0.50, 0.90, 0.975),
    names = FALSE
  ))

  data.frame(
    estimand = labels,
    mean = colMeans(natural_draws),
    lower_95 = intervals[, 1],
    lower_80 = intervals[, 2],
    median = intervals[, 3],
    upper_80 = intervals[, 4],
    upper_95 = intervals[, 5],
    row.names = NULL
  )
}
```

## Data

`Raw_data.csv` is hosted in the study's [OSF Electronic Supplement](https://osf.io/qrbzj/) and is not distributed through this repository. Download it to `Analysis/Raw_data.csv`, or set the `RATSHARING_DATA` environment variable to its local path. The file contains 145 session-level observations from three focal rats across seven conditions. The value `-1` denotes an outcome that was not applicable or not recorded; it is never analyzed as a behavioral count.

```{r data_validation}

DATA_FILE <- Sys.getenv("RATSHARING_DATA", unset = "Raw_data.csv")
if (!file.exists(DATA_FILE)) {
  stop(paste0(
    "Data file not found. Download Raw_data.csv from https://osf.io/qrbzj/ ",
    "and save it in Analysis/, or set RATSHARING_DATA to its local path."
  ))
}

raw_data <- read_csv(DATA_FILE, show_col_types = FALSE)
required_columns <- c(
  "Rat", "Condition", "FoodAmmt", "Social", "Food",
  "Sharing", "TimesShared", "PelletsLeft"
)

if (!all(required_columns %in% names(raw_data))) {
  stop("Raw_data.csv does not contain the published analysis variables.")
}

analysis_data <- raw_data
analysis_data$Subject <- as.integer(as.factor(analysis_data$Rat))

stopifnot(
  nrow(analysis_data) == 145,
  length(unique(analysis_data$Subject)) == 3,
  identical(as.integer(sort(unique(analysis_data$Condition))), 1:7),
  all(analysis_data$Food >= 0),
  all(analysis_data$FoodAmmt > 0),
  all(analysis_data$Social[analysis_data$Condition == 5] == -1),
  all(analysis_data$Social[analysis_data$Condition != 5] >= 0),
  all(analysis_data$Sharing[analysis_data$Condition == 5] == -1),
  all(analysis_data$Sharing[analysis_data$Condition != 5] >= 0),
  all(analysis_data$PelletsLeft[analysis_data$Condition %in% c(5, 7)] >= 0),
  all(analysis_data$PelletsLeft[!analysis_data$Condition %in% c(5, 7)] == -1)
)

expected_sessions <- matrix(
  c(
    11, 9, 13, 5, 5, 6, 6,
    11, 9,  9, 5, 5, 6, 7,
     5, 4,  5, 8, 5, 6, 5
  ),
  nrow = 3,
  byrow = TRUE
)
observed_sessions <- xtabs(~ Rat + Condition, data = analysis_data)
stopifnot(all(as.integer(observed_sessions) == as.integer(expected_sessions)))

social_for_summary <- replace(
  analysis_data$Social,
  analysis_data$Social < 0,
  NA_integer_
)
sharing_for_summary <- replace(
  analysis_data$Sharing,
  analysis_data$Sharing < 0,
  NA_integer_
)

data_profile <- data.frame(
  condition = 1:7,
  sessions = as.integer(table(analysis_data$Condition)),
  mean_food_choices = as.numeric(tapply(analysis_data$Food, analysis_data$Condition, mean)),
  mean_social_choices = as.numeric(tapply(social_for_summary, analysis_data$Condition, mean, na.rm = TRUE)),
  mean_shared_pellets = as.numeric(tapply(sharing_for_summary, analysis_data$Condition, mean, na.rm = TRUE))
)

cat("CmdStan:", INSTALLED_CMDSTAN_VERSION, "\n")
cat("Seeds:", paste(names(MODEL_SEEDS), MODEL_SEEDS, sep = "=", collapse = ", "), "\n")
data_profile
```

## Model 1: Food versus social choice

The published choice analysis excludes Condition 5 because the social option was unavailable. The remaining conditions are re-indexed consecutively for Stan. The model estimates condition-level food-choice log odds and correlated subject deviations.

```{r run_choice_model}
choice_rows <- analysis_data$Condition != 5
choice_subset <- analysis_data[choice_rows, ]
choice_subset$Condition_Stan <- match(
  choice_subset$Condition,
  sort(unique(choice_subset$Condition))
)

stan_data_choice <- list(
  N = nrow(choice_subset),
  S = length(unique(choice_subset$Subject)),
  subID = as.integer(choice_subset$Subject),
  C = length(unique(choice_subset$Condition_Stan)),
  condID = as.integer(choice_subset$Condition_Stan),
  choice = as.integer(choice_subset$Social + choice_subset$Food),
  food = as.integer(choice_subset$Food)
)

choice_model <- cmdstan_model("pair_multilevel_model.stan")
choice_fit <- sample_maintained_model(
  choice_model,
  stan_data_choice,
  MODEL_SEEDS[["choice"]]
)

choice_probability_summary <- summarize_gamma(
  choice_fit,
  paste("Condition", c(1:4, 6:7)),
  plogis
)
choice_probability_summary
choice_diagnostics <- report_diagnostics(choice_fit, "Choice model")
```

## Model 2: Food and social response rates

The response-rate analysis models food and social response counts jointly with a negative-binomial likelihood. Condition 5 remains in the model; its social count is `-1`, so that likelihood contribution is skipped exactly as in the published Stan program. The Condition 5 social-rate parameter therefore has no direct likelihood contribution and should not be interpreted as an empirical estimate.

```{r run_response_rate_model}
stan_data_response <- list(
  N = nrow(analysis_data),
  S = length(unique(analysis_data$Subject)),
  subID = as.integer(analysis_data$Subject),
  C = length(unique(analysis_data$Condition)),
  condID = as.integer(analysis_data$Condition),
  FC = as.integer(analysis_data$Food),
  SC = as.integer(analysis_data$Social)
)

response_model <- cmdstan_model("resp_rate_multilevel_model.stan")
response_fit <- sample_maintained_model(
  response_model,
  stan_data_response,
  MODEL_SEEDS[["response_rate"]]
)

response_rate_labels <- c(
  paste("Food | Condition", 1:7),
  paste("Social | Condition", 1:7)
)
response_rate_summary <- summarize_gamma(
  response_fit,
  response_rate_labels,
  exp
)
response_rate_summary$evidence <- ifelse(
  response_rate_summary$estimand == "Social | Condition 5",
  "prior only",
  "observed"
)
response_rate_summary
response_diagnostics <- report_diagnostics(response_fit, "Response-rate model")
```

## Model 3: Pellet allocation

The published pellet model has three outcomes per condition:

- `CP`: the published consumed-pellet measure, calculated as `Food * FoodAmmt` in the archived analysis.
- `LP`: pellets left behind, from `PelletsLeft`.
- `SP`: pellets shared, from `Sharing`.

Values of `-1` mark outcomes that were not applicable; the Stan model skips those likelihood contributions.

```{r run_pellet_model}
stan_data_pellet <- list(
  N = nrow(analysis_data),
  S = length(unique(analysis_data$Subject)),
  subID = as.integer(analysis_data$Subject),
  C = length(unique(analysis_data$Condition)),
  condID = as.integer(analysis_data$Condition),
  CP = as.integer(analysis_data$Food * analysis_data$FoodAmmt),
  LP = as.integer(analysis_data$PelletsLeft),
  SP = as.integer(analysis_data$Sharing)
)

stopifnot(
  identical(stan_data_pellet$CP, as.integer(analysis_data$Food * analysis_data$FoodAmmt)),
  identical(stan_data_pellet$LP, as.integer(analysis_data$PelletsLeft)),
  identical(stan_data_pellet$SP, as.integer(analysis_data$Sharing))
)

pellet_model <- cmdstan_model("intake_multilevel_model.stan")
pellet_fit <- sample_maintained_model(
  pellet_model,
  stan_data_pellet,
  MODEL_SEEDS[["pellet"]]
)

pellet_labels <- c(
  paste("Published CP | Condition", 1:7),
  paste("Left behind | Condition", 1:7),
  paste("Shared | Condition", 1:7)
)
pellet_rate_summary <- summarize_gamma(
  pellet_fit,
  pellet_labels,
  exp
)
pellet_prior_only <- (
  grepl("^Left behind", pellet_rate_summary$estimand) &
    !grepl("Condition (5|7)$", pellet_rate_summary$estimand)
) | pellet_rate_summary$estimand == "Shared | Condition 5"
pellet_rate_summary$evidence <- ifelse(
  pellet_prior_only,
  "prior only",
  "observed"
)
pellet_rate_summary
pellet_diagnostics <- report_diagnostics(pellet_fit, "Pellet model")
```

## Interpretation

The natural-scale tables report posterior means, medians, 80% credible intervals, and 95% credible intervals, matching the article's emphasis on estimated behavior and uncertainty. These tables apply the inverse link to `gamma`, so they are conditional population-level estimates at zero subject deviation rather than marginal averages across subjects. Parameters marked `prior only` lack a direct likelihood contribution and are retained solely to preserve the published model structure.

The substantive benchmark is the peer-reviewed article: food preference changed with food magnitude and home-cage food access, social responding remained comparatively stable, and sharing was rare across conditions. Re-executed estimates should be compared with those reported patterns and intervals rather than with the exact archived Monte Carlo draws.

## Reproducibility record

```{r reproducibility_record}
list(
  analysis_seeds = MODEL_SEEDS,
  cmdstan_version = as.character(cmdstan_version()),
  cmdstanr_version = as.character(packageVersion("cmdstanr")),
  r_version = R.version.string,
  platform = R.version$platform
)

sessionInfo()
```
